Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-10 brms_meeg.tex
updating the manuscript (intro)

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- or electro-encephalography (M/EEG) offer a
unique window into the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time, followed by corrections for multiple comparisons (e.g.,
FDR, FWER) or cluster-based inference. While effective for controlling
error rates, cluster-based methods come with a significant limitation:
they shift the focus of inference from individual time points to
clusters, making it difficult to draw precise conclusions about the
onset or offset of observed effects. Here, we introduce a
\emph{model-based} approch for analysing M/EEG timeseries such as
event-related potentials (ERPs) or decoding performance over time. Our
method leverages Bayesian generalised additive multilevel models
(GAMMs), providing posterior probabilities that an effect is above zero
(or above chance) at each time point, while naturally accounting for
temporal dependencies and between-subject variability. Using both
simulated and actual M/EEG datasets, we demonstrate that this approach
substantially outperforms conventional methods in estimating the onset
and offset of neural effects, yielding more precise and reliable
results. We also provide an R package implementing the method and show
how it can be seamlessly integrated into M/EEG analysis pipelines using
MNE-Python.}

\keywords{EEG, MEG, generalised additive models, mixed-effects
models, multilevel models, Bayesian statistics, brms}

@@

\usepackage{mathtools}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography (MEG) and
electroencephalography (EEG) are two such methods, widely used in
cognitive neuroscience for their ability to track brain activity at the
millisecond scale. These techniques provide rich time series data that
reflect how neural responses unfold in response to stimuli or tasks. A
central goal in many M/EEG studies is to determine whether and when
neural responses differ across experimental conditions or participant
groups.

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pernet2015}{Pernet et al., 2015})\ldots{} See also Maris
(\citeproc{ref-maris2011}{2011}) and Rosenblatt et al.
(\citeproc{ref-rosenblatt2018}{2018}) (history of cluster-based
approaches)\ldots{} Clusters failures (\citeproc{ref-eklund2016}{Eklund
et al., 2016})\ldots{}

@@

Description of cluster-based approaches (see
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}

However, as pointed by the original authors themselves
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}), ``there is a
conflict between this interest in localized effects and our choice for a
global null hypothesis: by controlling the FA {[}false alarm{]} rate
under this global null hypothesis, one cannot quantify the uncertainty
in the spatiotemporal localization of the effect''
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). In
contrast, the proposed approach, based on Bayesian generalised additive
multilevel models, allows quantifying the posterior probability of
effects being above chance at the level of timesteps, voxels, sensors
(etc), while naturally taking into account spatiotemporal dependencies
present in M/EEG timeseries.

\subsection{Related work}\label{related-work}

@@

@@

Recently, Teichmann (\citeproc{ref-teichmann2022}{2022}) provided a
detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D
output from MVPA, that is, for testing, at every timestep, whether
decoding performance is above chance level. However, this approach
provides timeseries of BFs that ignores temporal dependencies..

\subsection{Bayesian regression
modelling}\label{bayesian-regression-modelling}

Short intro/recap about Bayesian (linear and generalised) regression
models\ldots{}

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this stage: \emph{splines} are functions composed of
simpler functions. These simpler functions are basis functions (e.g., a
cubic polynomial) and the set of basis functions is a \emph{basis}. Each
basis function gets its coefficient and the resultant spline is the sum
of these weighted basis functions. Splines coefficients are penalised
(usually through the squared of the smooth functions' second derivative)
in a way that can be interpreted, in Bayesian terms, as a prior on the
``wiggliness'' of the function. In other words, more complex (wiggly)
basis function are penalised.

@@

Introduction to multilevel GAMs
(\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al.,
2019})\ldots{}

Now describe the Bayesian GAMM (\citeproc{ref-miller2025}{Miller,
2025})\ldots{} Proper inclusion of varying/random effects in the model
specification protects against overly wiggly curves
(\citeproc{ref-baayen2020}{Baayen \& Linke, 2020})\ldots{} Generalising
to scale and shape or ``distributional GAMs''
(\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}) and applied to
neuroimaging data (\citeproc{ref-dinga2021}{Dinga et al., 2021})\ldots{}

@@

Given the previously reported limitations of conventional methods to
precisely identify the onset and offset of M/EEG effects (e.g., ERPs,
decoding performance), we aimed to provide a model-based approach for
estimating the onset and offset of such effects. To achieve this, we
leveraged Bayesian generalised additive multilevel models (BGAMMs)
fitted via the \texttt{brms\ R} package and compared the performance of
this approach to conventional methods on both simulated and actual M/EEG
data.

@@

Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see code in the online supplementary materials and
details in \citeproc{ref-yeung2004}{Yeung et al., 2004}). As in
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), the signal was
generated from truncated Gaussian with an objective onset at 160 ms, a
peak at 250 ms, and an offset at 342 ms. We simulated this signal for
250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We
simulated data for a group of 20 participants with 50 trials per
participant and condition (Figure~\ref{fig-eeg}).

@@

\subsection{Model fitting}\label{model-fitting}

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\) (\citeproc{ref-gabry2019}{Gabry et al., 2019};
\citeproc{ref-gelman2020}{Gelman et al., 2020}). The \texttt{brms}
package uses the same syntax as the \texttt{R} package \texttt{mgcv} v
1.9-1 (\citeproc{ref-mgcv}{Wood, 2017b}) for specifying smooth effects.
Figure~\ref{fig-plot-post-slope} shows the predictions of this model
together with the raw data.

@@

\CommentTok{\# timing the model fitting}
\CommentTok{\# around 600s for 2000 iterations on 4 chains and 4 cores}
\CommentTok{\# around 190s for 2000 iterations on 4 chains and 4 cores + cmdstanr and 01}
\CommentTok{\# around 440s for 2000 iterations on 4 chains and 4 cores + threading(2), cmdstanr and 01}
\CommentTok{\# library(tictoc)}
\CommentTok{\# tic()}

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
    \CommentTok{\# eeg \textasciitilde{} condition + s(time, bs = "cr", k = 20, by = condition),}
    \CommentTok{\# thin{-}plate regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
    \DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
    \CommentTok{\# speeding up the model fitting \#}
    \DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
    \CommentTok{\# backend = "cmdstanr",}
    \CommentTok{\# threads = threading(2),}
    \CommentTok{\# stan\_model\_args = list(stanc\_options = list("O1") )}
\NormalTok{    )}

\CommentTok{\# stop timing}
\CommentTok{\# toc()}
\end{Highlighting}
\end{Shaded}

We depict the posterior predictions together with the posterior estimate
of the slope for \texttt{condition} at each timestep
(Figure~\ref{fig-plot-post-slope}).

@@

\caption{\label{fig-plot-post-slope}Posterior estimate of the EEG
activity in each condition (left) and posterior estimate of the
difference in EEG activity (right) according to the GAM.}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0\) (Figure~\ref{fig-post-prob-ratio},
left). We can also express this as the ratio of posterior probabilities
(i.e., \(p/(1-p)\)) and visualise the timecourse of this ratio
superimposed with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right). A ratio of 10 means that the
probability of the difference being above 0 is 10 times higher than the
probability of the difference not being above 0, given the data, the
priors, and other model's assumptions.

@@

\caption{\label{fig-post-prob-ratio}Left: Posterior probability of the
EEG difference (slope) being above 0 according to the GAM. Right: Ratio
of posterior probability according to the GAM (on a log10 scale).
Timesteps above threshold (10) are highlighted in green.}

@@

\subsection{Multilevel modelling using summary
statistics}\label{multilevel-modelling-using-summary-statistics}

The previous model only included constant (fixed) effects, thus not
properly accounting for between-participant variability. We next fit a
multilevel version of the GAM. Although it is possible to fit a GAMM at
the single-trial level, we present a resource-efficient version of the
model that is fitted directly on the by-participant summary statistics
(mean and SD), similar to what is done in meta-analysis.

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

>>>>

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- or electro-encephalography (M/EEG) offer a
unique window into the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time and space, followed by corrections for multiple comparisons
such as cluster-based inference. While effective for controlling error
rates at the cluster-level, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it makes difficult to draw precise
conclusions about the onset or offset of observed effects. Here, we
introduce a \emph{model-based} approach for analysing M/EEG timeseries
such as event-related potentials (ERPs) or decoding performance over
time. Our method leverages Bayesian generalised additive multilevel
models (GAMMs), providing posterior probabilities that an effect is
above zero (or above chance) at each time point, while naturally
accounting for temporal dependencies and between-subject variability.
Using both simulated and actual M/EEG datasets, we demonstrate that this
approach substantially outperforms conventional methods in estimating
the onset and offset of neural effects, yielding more precise and
reliable results. We provide an R package implementing the method and
show how it can be seamlessly integrated into M/EEG analysis pipelines
using MNE-Python.}

\keywords{EEG, MEG, cluster-based inference, simulation, multiple
comparisons, generalised additive models, mixed-effects
models, multilevel models, Bayesian statistics, brms}

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage{booktabs, caption, longtable, colortbl, array}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

@@

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography and electroencephalography
(M/EEG) are two such methods, widely used in cognitive neuroscience for
their ability to track brain activity at the millisecond scale. These
techniques provide rich time series data that reflect how neural
responses unfold in response to stimuli or tasks. A central goal in many
M/EEG studies is to determine whether, when, and where neural responses
differ across experimental conditions or participant groups.

The conventional approach involves mass-univariate statistical testing
with a unique threshold, followed by corrections for multiple
comparisons with the goal of maintaining the familywise error rate
(FWER) or false discovery rate (FDR) at the nominal level. Cluster-based
inference is the most common way of achieving this sort of error control
in the M/EEG literature, being the default option in several software
programs. While effective for controlling error rates, cluster-based
inference comes with a significant limitation: they shift the focus of
inference from individual time points to clusters, making it difficult
to draw precise conclusions about the onset or offset of observed
effects (\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}); ``there
is a conflict between this interest in localized effects and our choice
for a global null hypothesis: by controlling the FA {[}false alarm{]}
rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, as noted by Rosenblatt et al.
(\citeproc{ref-rosenblatt2018}{2018}), cluster-based inference suffers
from low spatial resolution: ``Since discovering a cluster means that
`there exists at least one voxel with an evoked response in the
cluster', and not that `all the voxels in the cluster have an evoked
response', it follows that the larger the detected cluster, the less
information we have on the location of the activation.'' As a
consequence, cluster-based inferenc has been shown\ldots{}

To overcome the limitations of cluster-based inference, we introduce a
novel model-based approach for precisely localising M/EEG effects in
time (space, and other dimensions). The proposed approach, based on
Bayesian generalised additive multilevel models, allows quantifying the
posterior probability of effects being above chance at the level of
timesteps, sensors, voxels, etc., while naturally taking into account
spatiotemporal dependencies present in M/EEG timeseries. We compare the
performance of the proposed approach to well-established alternative
methods using both simulated and actual M/EEG data and show that it
significantly outperforms alternative methods in estimating the onset
and offset of M/EEG effects.

\subsection{Cluster-based inference}\label{cluster-based-inference}

Different methods exist to control the family-wiser error rate (FWER),
defined as the type-1 error rate (false positive) over an ensemble
(family) of tests\ldots{} for instance the Bonferroni correction
(\citeproc{ref-dunn1961}{Dunn, 1961}), however this method is generally
overconservative as it assumes statistical independence of tests, an
assumptions that is clearly violated in the context of M/EEG timeseries
with spatiotemporal dependencies\ldots{} or the false discovery rate
(FDR), defined as the proportion of false positive among positive tests
(e.g., \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001})\ldots{}

On popular technique to account for spatiotemporal dependencies while
controlling the FWER is cluster-based inference. Description of
cluster-based approaches (see \citeproc{ref-sassenhagen2019}{Sassenhagen
\& Draschkow, 2019})\ldots{} A typical cluster-based inference consists
of two successive steps. First, clusters are defined as sets of
contiguous voxels, channels, and/or timesteps, whose intensity/activity
exceeds some predefined threshold. Clusters are then characterised by
their height (i.e., maximal value), extent (number of constituent
elements), or some combination of both, such at its ``mass'', for
instance by summing the statistics within a cluster, an approach
refereed to as ``cluster mass'' (\citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-pernet2015}{Pernet et al., 2015}).
Then, the null hypothesis is tested by assessing the probability.. As
different cluster-forming thresholds lead to clusters with different
spatial or temporal extent, this initial threshold modulates the
sensitivity of the subsequent permutation test. The threshold-free
cluster enhancement method (TFCE) was introduced by S. Smith \& Nichols
(\citeproc{ref-smith2009}{2009}) to overcome this choice of an arbitrary
threshold.

In brief, the TFCE method works as follows. Instead of picking an
arbitrary cluster-forming threshold (e.g., \(t=2\)), we try all (or
many) possible thresholds in a given range and check whether a given
timestep/voxel belongs to a significant cluster under any of the set of
thresholds. Then, instead of using cluster mass (e.g., the sum of
squared t-values within the cluster), we use a weighted average between
the cluster extend (\(e\), how broad is the cluster, that is, how many
connected samples it contains) and the cluster height (\(h\), how high
is the cluster, that is, how large is the test statistic). The TFCE
score at each timestep/voxel \(t\) is given by:

\[
\text{TFCE(t)} = \int_{h=h_{0}}^{h_{t}} e(h)^{E} h^{H} \mathrm{d}h
\]

where \(h_{0}\) is typically \(0\) and parameters \(E\) and \(H\) are
set a priori (typically to 0.5 and 2, respectively) and control the
influence of the extend and height on the TFCE. Then, p-value for
timestep/voxel \(t\) is computed by comparing it TFCE with the null
distribution of TFCE values. For each permuted signal, we keep the
maximal value over the whole signal for the null distribution of the
TFCE. The TFCE combined with permutation (assuming a large enough number
of permutations) has been shown to provide accurate type 1 FWER (e.g.,
\citeproc{ref-pernet2015}{Pernet et al., 2015}). However, previous
simulation work showed that cluster-based methods have indeed poor
performance in localising the onset of M/EEG effects (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To sum up this section, cluster-based inference main limitations is that
it provides inference at the cluster level only, not allowing inference
at the level of timesteps, sensors, etc. Thus, it does not allow
inferring the localisation of effects (\citeproc{ref-maris2011}{Maris,
2011}; for more details on cluster-based inference, see for instance
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). In the
following, we briefly review previous modelling work. Then, we provide a
brief introduction to Bayesian statistical modelling, before moving to
generalised additive models (GAMs) and generalised additive multilevel
models (GAMMs) to illustrate how they can be used to precisely localise
the onset and offset of M/EEG effects.

\subsection{Previous work on modelling M/EEG
data}\label{previous-work-on-modelling-meeg-data}

@@

As put by Rousselet (\citeproc{ref-rousselet_using_2025}{2025}),
concluding on the onset of effect based on a series of univariate
tests\ldots{} commits to three fallacies\ldots{} here, we want to avoid
these by introducing a model-based approach, which naturally take into
account the temporal dependencies in the data to output a series of
posterior probabilities\ldots{}

@@

Recently, Teichmann (\citeproc{ref-teichmann2022}{2022}) provided a
detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D
output from MVPA, that is, for testing, at every timestep, whether
decoding performance is above chance level. However, this approach
provides timeseries of BFs that ignores temporal dependencies\ldots{}

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this stage: \emph{splines} are functions composed of
simpler functions. These simpler functions are basis functions (e.g.,
cubic polynomial, thin-plate) and the set of basis functions is a
\emph{basis}. Each basis function gets its coefficient and the resultant
spline is the sum of these weighted basis functions
(Figure~\ref{fig-intro-gam}). Splines coefficients are penalised
(usually through the squared of the smooth functions' second derivative)
in a way that can be interpreted, in Bayesian terms, as a prior on the
``wiggliness'' of the function. In other words, more complex (wiggly)
basis function are penalised.

@@

The Bayesian approach to statistical modelling is characterised by its
reliance on probability theory to (\citeproc{ref-gelman2020}{Gelman et
al., 2020})\ldots{} In this framework, all unknown entities are assigned
probability distributions reflecting the uncertainty\ldots{} These
probability distributions are commonly referred to as ``priors'' and
represent some state of knowledge about unknown quantities before seeing
any data. There are debates among Bayesian practitioners as to whether
prior distributions should encore subjective (personal) beliefs
or\ldots{} but these debates are outside the scope of the present paper
and we therefore the interested reader to dedicated work (e.g., XX;
YY)\ldots{} In practice, weakly informative priors are often used as
default priors in situations in which subjective priors are difficult to
define/elicit\ldots{} Bayesian models are then fitted on empirical
(actual or simulated) data to update prior states of knowledge to
posterior states of knowledge using Bayes theorem, or in practise,
sampling-based approximations of the posterior distribution\ldots{}

\caption{\label{fig-intro-gam}Different types of GAM(M)s. \textbf{A}:
GAMs predictions are computed as the weigthed sum (in black) of basis
functions (here thin-plate basis functions, in grey). \textbf{B}:
Constant-effect GAM, with 5 participants in colours and the group-level
prediction in black. \textbf{C}: Varying-intercept + varying-slope GAMM
(with common smoother). \textbf{D}: Varying-intercept + varying-slope +
varying-smoother GAMM. In this model, each participant gets its own
intercept, slope, and degree of `wiggliness' (smoother).}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-intro-gam-1.png}

See Figure~\ref{fig-intro-gam}\ldots{} Introduction to multilevel GAMs
(\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al.,
2019})\ldots{} Now describe the Bayesian GAMM
(\citeproc{ref-miller2025}{Miller, 2025})\ldots{} Proper inclusion of
varying/random effects in the model specification protects against
overly wiggly curves (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020})\ldots{} Generalising to scale and shape or ``distributional
GAMs'' (\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}) and applied to
neuroimaging data (\citeproc{ref-dinga2021}{Dinga et al., 2021})\ldots{}

@@

Given the previously reported limitations of conventional methods to
precisely identify the onset and offset of M/EEG effects (e.g., ERPs,
decoding performance), we developed a model-based approach for
estimating the onset and offset of such effects. To achieve this, we
leveraged Bayesian generalised additive multilevel models (BGAMMs)
fitted in \texttt{R} via the \texttt{brms} package and compared the
performance of this approach to conventional methods on both simulated
and actual M/EEG data.

@@

Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see code in the online supplementary materials and
details in \citeproc{ref-yeung2004}{Yeung et al., 2004}). As in
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), the signal was
generated from a truncated Gaussian distribution with an objective onset
at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this
signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling
rate. We simulated data for a group of 20 participants (with variable
true onset) with 50 trials per participant and condition
(Figure~\ref{fig-eeg}).

@@

\subsection{Model description and model
fitting}\label{model-description-and-model-fitting}

We then fitted a Bayesian GAM (BGAM) using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\) (\citeproc{ref-gabry2019}{Gabry et al., 2019};
\citeproc{ref-gelman2020}{Gelman et al., 2020}). The \texttt{brms}
package uses the same syntax as the \texttt{R} package \texttt{mgcv} v
1.9-1 (\citeproc{ref-mgcv}{Wood, 2017b}) for specifying smooth effects.
Figure~\ref{fig-plot-post-slope} shows the predictions of this model
together with the raw data.

@@

\CommentTok{\# fitting the BGAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
    \CommentTok{\# eeg \textasciitilde{} condition + s(time, bs = "cr", k = 20, by = condition),}
    \CommentTok{\# thin{-}plate regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

However, the previous model only included constant (fixed) effects, thus
not properly accounting for between-participant variability. We next fit
a multilevel version of the BGAM (BGAMM). Although it is possible to fit
a BGAMM at the single-trial level, we present a resource-efficient
version of the model that is fitted directly on the by-participant
summary statistics (mean and SD), similar to what is done in
meta-analysis.

\CommentTok{\# fitting the BGAMM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

We depict the posterior predictions together with the posterior estimate
of the slope for \texttt{condition} at each timestep
(Figure~\ref{fig-plot-post-slope}). This figure suggests that the BGAMM
provides an adequate description of the simulated data.

@@

\caption{\label{fig-plot-post-slope}Posterior estimate of the EEG
activity in each condition (left) and posterior estimate of the
difference in EEG activity (right) according to the BGAMM.}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0\) (Figure~\ref{fig-post-prob-ratio},
left). We can also express this as the ratio of posterior probabilities
(i.e., \(p/(1-p)\)) and visualise the timecourse of this ratio
superimposed with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right). Note that a ratio of 10 means
that the probability of the difference being above 0 is 10 times higher
than the probability of the difference not being above 0, given the
data, the priors, and other model's assumptions. Thresholding the
posterior probability ratio thus provides a model-based approach for
estimating the onset and offset of M/EEG effects.

@@

\caption{\label{fig-post-prob-ratio}Left: Posterior probability of the
EEG difference (slope) being above 0 according to the BGAMM. Right:
Ratio of posterior probability according to the BGAMM (on a log10
scale). Timesteps above threshold (10) are highlighted in green.}

@@

@@

The TFCE performs worse than the cluster-sum approach, which was
anticipated by Rousselet (\citeproc{ref-rousselet_using_2025}{2025})
based on the initial results of S. Smith \& Nichols
(\citeproc{ref-smith2009}{2009})\ldots{}
